Rodent trapping studies as an overlooked information source for understanding endemic and novel zoonotic spillover.

### 2022-08-25 reports/manuscript_v5.Rmd
version for archiving

---
title: "Rodent trapping studies as an overlooked information source for understanding endemic and novel zoonotic spillover."
date: "`r Sys.Date()`"
author: 
  - "David Simons"
  - "Lauren A. Attfield"
  - "Kate E. Jones"
  - "Deborah Watson-Jones"
  - "Richard Kock"
output: 
  officedown::rdocx_document:
    tables:
      caption:
       style: false
    plots:
      caption:
        style: false
  # pdf_document
always_allow_html: yes
bibliography: citations/manuscript_citations.bib
csl: harvard.csl
latex_engine: xelatex
---

@@

![Fig 1: A) The location of trapping sites in West Africa. No sites were recorded from Togo or The Gambia. Heterogeneity is observed in the coverage of each country by trap night (colour) and location of sites. For example, Senegal, Mali and Sierra Leone have generally good coverage compared to Guinea and Burkina Faso. B) Histogram of trap nights performed at each study site, a median of 248 trap nights (IQR 116-500) was performed at each site](`r here("figures", "Figure_1_updated.pdf")`)

@@

To investigate the extent of spatial bias in the rodent trapping data, we calculated trap-night (TN) density within each West African level-2 administrative region. The `sf` package in the R statistical language was used to manipulate geographic data, administrative boundaries were obtained from GADM 4.0.4 [@pebesma_simple_2018; @r_core_team_r_2020; @database_of_global_administrative_areas_gadm_2022].  Density ($\text{TN}_\text{density}$) was calculated by dividing the number of trap nights by the area of a level-2 administrative area ($\text{R}_\text{area}$). For studies not reporting trap nights imputation was used as previously described. Human population density was obtained for the closest year (2005) to the median year of trapping (2007) from Socioeconomic Data and Applications Center (SEDAC) gridded population of the world v4 at ~ 1km resolution ($\text{P}_\text{density}$) [@socioeconomic_data_and_applications_center_gridded_2021].  Median population density was then calculated for each level-2 administrative region. Land cover classification was obtained from the Copernicus climate change service at ~300m resolution [@european_space_agency_climate_change_initiative_land_2022]. The proportion of cropland, shrubland, tree cover (${\psi}_\text{tree}$) and urban land cover (${\psi}_\text{urban}$) within a level-2 administrative region in 2005 was calculated.

@@

$$
\text{TN}_\text{density} \sim \text{Tweedie}(\text{Pop}_\text{density} + \text{R}_\text{area} + {\psi}_\text{urban} + (X * Y))
$$

@@

We found a non-random spatial distribution of rodent trapping locations across the study region, suggestive of underlying bias in the sampling or rodents across West Africa. Trap sites were situated in 256 of 1,450 (17.6%) level-2 administrative regions in 14 West African nations. The regions with highest TN density included the capitals and large cities of Niger (Niamey), Nigeria (Ibadan), Ghana (Accra), Senegal (Dakar), Ghana (Accra) and Benin (Cotonou). Outside of these cities, regions in, Northern Senegal, Southern Guinea, Edo and Ogun States in Nigeria and Eastern Sierra Leone had the greatest TN density (Fig 1A.).

@@

![Fig 2. Relative trapping effort bias across West Africa from included studies adjusted for human population density, proportion urban land cover and area of the administrative region. Yellow regions represent areas with a bias towards increased trapping effort (e.g., North West Senegal), purple regions represent areas with a bias towards reduced trapping effort (e.g., Northern Nigeria).](`r here("figures", "Figure_2_updated.pdf")`)

@@

![Fig 3. Each row corresponds to a single rodent species. L) Presence recorded in GBIF overlaid on IUCN species range (red-shaded area). R) Detection and non-detection from rodent trapping studies overlaid on IUCN species ranges. M. musculus has no IUCN West African range.](`r here("figures", "Figure_3_updated.pdf")`)

@@

![Fig 4. A) Identified species level host-pathogen associations through detection of acute infection (i.e. PCR, culture). Percentages and colour relate to the proportion of all assays that were positive, the number of individuals tested for the pathogen is labelled N. Associations with a black border are present in the CLOVER dataset.](`r here("figures", "Figure_4a_updated.pdf")`)

![Fig 4. B) Identified species level host-pathogen associations through serological assays (i.e. ELISA). Percentages and colour relate to the proportion of all assays that were positive, the number of individuals tested for the pathogen is labelled N. Associations with a black border are present in the CLOVER dataset.](`r here("figures", "Figure_4b_updated.pdf")`)

@@

![Supplementary Fig 1. Green points represent the start date of rodent trapping studies, blue points representing the final trapping activity. Red points indicate the publication of studies. Increasing numbers of studies have been published since 2000 with more studies being conducted over repeated visits.](`r here("figures", "Supplementary_Figure_1.pdf")`)

@@

![Supplementary Fig 2.](`r here("figures", "Supplementary_Figure_2.pdf")`)

@@

![Supplementary Fig 3A. Relative trapping effort bias across West Africa from the subset of included studies reporting trapping effort adjusted for proportion urban land cover and proportion tree cover. Predictions are limited to areas around trap sites (coloured areas), uncertainty in the estimate of bias is represented by colour transparency/pallor. Purple regions represent areas with higher than expected trapping effort, yellow regions represent areas lower than expected trapping effort](`r here("figures", "Figure_2_sensitivity.png")`)

![](`r here("figures", "Figure_2_ss.png")`)

## Supplementary Fig 4
![](`r here("figures", "Supplementary_Figure_4.pdf")`)

![Supplementary Fig 4. A) Identified host-pathogen associations at pathogen family level through detection of acute infection (i.e. PCR, culture). Percentages and colour relate to the proportion of all assays that were positive. Associations with a black border are present in the CLOVER dataset. B) Identified host-pathogen associations at pathogen family level through serological assays (i.e. ELISA). Percentages and colour relate to the proportion of all assays that were positive. Associations with a black border are present in the CLOVER dataset.](`r here("figures", "Supp_4b_updated.png")`)

>>>>

---
title: "Rodent trapping studies as an overlooked information source for understanding endemic and novel zoonotic spillover."
date: "`r Sys.Date()`"
author: 
  - "David Simons"
  - "Lauren A. Attfield"
  - "Kate E. Jones"
  - "Deborah Watson-Jones"
  - "Richard Kock"
output: 
  officedown::rdocx_document:
    tables:
      caption:
       style: false
    plots:
      caption:
        style: false
  # pdf_document
always_allow_html: yes
bibliography: citations/manuscript_citations.bib
csl: plos.csl
latex_engine: xelatex
---

@@

![Fig 1: A) The location of trapping sites in West Africa. No sites were recorded from Togo or The Gambia. Heterogeneity is observed in the coverage of each country by trap night (colour) and location of sites. For example, Senegal, Mali and Sierra Leone have generally good coverage compared to Guinea and Burkina Faso. B) Histogram of trap nights performed at each study site, a median of 248 trap nights (IQR 116-500) was performed at each site](`r here("figures", "Figure_1_updated.png")`)

@@

To investigate the extent of spatial bias in the rodent trapping data, we calculated trap-night (TN) density within each West African level-2 administrative region. The `sf` package in the R statistical language (R version 4.1.2) was used to manipulate geographic data, administrative boundaries were obtained from GADM 4.0.4 [@pebesma_simple_2018; @r_core_team_r_2020; @database_of_global_administrative_areas_gadm_2022].  Density ($\text{TN}_\text{density}$) was calculated by dividing the number of trap nights by the area of a level-2 administrative area ($\text{R}_\text{area}$). For studies not reporting trap nights imputation was used as previously described. Human population density was obtained for the closest year (2005) to the median year of trapping (2007) from Socioeconomic Data and Applications Center (SEDAC) gridded population of the world v4 at ~ 1km resolution ($\text{P}_\text{density}$) [@socioeconomic_data_and_applications_center_gridded_2021].  Median population density was then calculated for each level-2 administrative region. Land cover classification was obtained from the Copernicus climate change service at ~300m resolution [@european_space_agency_climate_change_initiative_land_2022]. The proportion of cropland, shrubland, tree cover (${\psi}_\text{tree}$) and urban land cover (${\psi}_\text{urban}$) within a level-2 administrative region in 2005 was calculated.

@@

$$
\text{TN}_\text{density} \sim \text{Tweedie}(\text{P}_\text{density} + \text{R}_\text{area} + {\psi}_\text{urban} + (X * Y))
$$

@@

We found non-random spatial clustering of rodent trapping locations across the study region, suggestive of underlying bias in the sampling or rodents across West Africa. Trap sites were situated in 256 of 1,450 (17.6%) level-2 administrative regions in 14 West African nations. The regions with highest TN density included the capitals and large cities of Niger (Niamey), Nigeria (Ibadan), Ghana (Accra), Senegal (Dakar), Ghana (Accra) and Benin (Cotonou). Outside of these cities, regions in, Northern Senegal, Southern Guinea, Edo and Ogun States in Nigeria and Eastern Sierra Leone had the greatest TN density (Fig 1A.).

@@

![Fig 2. Relative trapping effort bias across West Africa from included studies adjusted for human population density, proportion urban land cover and area of the administrative region. Yellow regions represent areas with a bias towards increased trapping effort (e.g., North West Senegal), purple regions represent areas with a bias towards reduced trapping effort (e.g., Northern Nigeria).](`r here("figures", "Figure_2_updated.png")`)

@@

![Fig 3. Each row corresponds to a single rodent species. L) Presence recorded in GBIF overlaid on IUCN species range (red-shaded area). R) Detection and non-detection from rodent trapping studies overlaid on IUCN species ranges. M. musculus has no IUCN West African range.](`r here("figures", "Figure_3_updated.png")`)

@@

![Fig 4) Identified species level host-pathogen associations through detection of acute infection (i.e. PCR, culture). Percentages and colour relate to the proportion of all assays that were positive, the number of individuals tested for the pathogen is labelled N. Associations with a black border are present in the CLOVER dataset.](`r here("figures", "Figure_4_updated.png")`)

![Fig 5) Identified species level host-pathogen associations through serological assays (i.e. ELISA). Percentages and colour relate to the proportion of all assays that were positive, the number of individuals tested for the pathogen is labelled N. Associations with a black border are present in the CLOVER dataset.](`r here("figures", "Figure_5_updated.png")`)

@@

![Supplementary Fig 1. Green points represent the start date of rodent trapping studies, blue points representing the final trapping activity. Red points indicate the publication of studies. Increasing numbers of studies have been published since 2000 with more studies being conducted over repeated visits.](`r here("figures", "Supplementary_Figure_1.png")`)

@@

![Supplementary Fig 2.](`r here("figures", "Supplementary_Figure_2.png")`)

@@

![Supplementary Fig 3A. Relative trapping effort bias across West Africa from the subset of included studies reporting trapping effort adjusted for proportion urban land cover and proportion tree cover. Predictions are limited to areas around trap sites (coloured areas), uncertainty in the estimate of bias is represented by colour transparency/pallor. Purple regions represent areas with higher than expected trapping effort, yellow regions represent areas lower than expected trapping effort](`r here("figures", "Supplementary_Figure_3.png")`)

## Supplementary Fig 4
![](`r here("figures", "Supplementary_Figure_4a.png")`)

![Supplementary Fig 4. A) Identified host-pathogen associations at pathogen family level through detection of acute infection (i.e. PCR, culture). Percentages and colour relate to the proportion of all assays that were positive. Associations with a black border are present in the CLOVER dataset. B) Identified host-pathogen associations at pathogen family level through serological assays (i.e. ELISA). Percentages and colour relate to the proportion of all assays that were positive. Associations with a black border are present in the CLOVER dataset.](`r here("figures", "Supplementary_Figure_4b.png")`)
